---
title: "ELMER: An R/Bioconductor Tool Inferring Regulatory Element Landscapes and Transcription Factor Networks Using Methylomes"
author: "Lijing Yao, Tiago Chedraoui Silva [ctb], Simon Coetzee [ctb], Ben Berman [aut], Peggy Farnham [aut]Hui Shen [ctb], Peter Laird [ctb]"
date: "`r Sys.Date()`"
output: 
  html_document:
    self_contained: true
    number_sections: no
    theme: flatly
    highlight: tango
    mathjax: null
    toc: true
    toc_float: true
    toc_depth: 2
    css: style.css
  
bibliography: bibliography.bib    
vignette: >
  %\VignetteIndexEntry{"ELMER: An R/Bioconductor Tool Inferring Regulatory Element Landscapes and Transcription Factor Networks Using Methylomes"}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


# Loading required data for plot
```{r,eval=TRUE, message=FALSE, warning = FALSE, results = "hide"}
# Load results from previous sections
mae <- get(load("mae.rda"))
```

# Scatter plots

##  Scatter plot of one probe and its nearby genes

Generate scatter plots for one probes' nearby 20 gene expression vs DNA methylation at this probe.

Each scatter plot shows the methylation level of an example  probe cg19403323 in all LUSC samples plotted against the expression of one of  20 adjacent genes.
```{r results='hide', echo=TRUE, message=FALSE, warning=FALSE, fig.height=5, fig.cap="Each scatter plot shows the methylation level of an example  probe cg19403323 in all LUSC samples plotted against the expression of one of  20 adjacent genes."}
scatter.plot(data = mae,
             byProbe = list(probe = c("cg19403323"), numFlankingGenes = 20), 
             category = "definition", 
             lm = TRUE, # Draw linear regression curve
             save = FALSE) 
```

##  Scatter plot of one pair
Generate a scatter plot for one probe-gene pair. Figure \ref{fig:figure2}

Scatter plot shows the methylation level of an example probe cg19403323 in all LUSC samples plotted against the expression of the putative  target gene SYT14.
```{r results='hide',eval=TRUE, fig.cap="Scatter plot shows the methylation level of an example probe cg19403323 in all LUSC samples plotted against the expression of the putative  target gene SYT14."}
scatter.plot(data = mae,
             byPair = list(probe = c("cg19403323"), gene = c("ENSG00000143469")), 
             category = "definition", save = TRUE, lm_line = TRUE) 
```

##  TF expression vs. average DNA methylation
Generate scatter plot for TF expression vs average DNA methylation of the sites 
with certain motif.

Each scatter plot shows the average  methylation level of sites with the TP53 motif in all LUSC samples plotted against the expression of the transcription factor TP53, TP63, TP73 respectively.

```{r,eval=TRUE, warning=FALSE, fig.cap="Each scatter plot shows the average  methylation level of sites with the TP53 motif in all LUSC samples plotted against the expression of the transcription factor TP53, TP63, TP73 respectively."}
load("result/getMotif.hypo.enriched.motifs.rda")
scatter.plot(data = mae,
             byTF = list(TF = c("TP53","TP63","TP73"),
                         probe = enriched.motif[["P73_HUMAN.H10MO.A"]]), 
             category = "definition",
             save = TRUE, 
             lm_line = TRUE)
```

