---
title: "4.4 - Regulatory TF plots"
author: "Lijing Yao [aut], Tiago Chedraoui Silva [aut], Simon Coetzee [aut], Ben Berman [aut], Houtan Noushmehr [ctb], Dennis J. Hazelett [ctb], Peggy Farnham [aut], Hui Shen [ctb], Peter Laird [ctb]"
date: "`r Sys.Date()`"
output: 
  html_document:
    self_contained: true
    number_sections: no
    theme: flatly
    highlight: tango
    mathjax: null
    toc: true
    toc_float: true
    toc_depth: 2
    css: style.css
  
bibliography: bibliography.bib    
vignette: >
  %\VignetteIndexEntry{"4.4 - Regulatory TF plots"}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


# TF ranking plot
TF ranking plot shows statistic $-log_{10}(P value)$ assessing the anti-correlation level 
of TFs expression level with average DNA methylation level at sites with a given motif. 
By default, the top 3 associated TFs and the TF family members (dots in red) that are associated with that specific motif  are labeled in the plot.
But there is also a option show highlight only TF sub-family members (TCClass database classification)

## TF ranking plot: family classification

Shown are TF ranking plots based on the score ($-log_{10}(P value))$ of association between TF expression and DNA methylation of the P73 motif in the LUSC cancer type. The dashed line indicates the boundary of the top 5% association score. The top 3 associated TFs and the TF family members=(dots in red) that are associated with that specific motif are labeled in the plot

```{r,eval=TRUE,fig.cap="Shown are TF ranking plots based on the score (-log(P value)) of association between TF expression and DNA methylation of the SOX2 motif in the LUSC cancer type. The dashed line indicates the boundary of the top 5% association score. The top 3 associated TFs and the TF family members=(dots in red) that are associated with that specific motif are labeled in the plot"}
load("result/getTF.hypo.TFs.with.motif.pvalue.rda")
motif <- "SOX2_HUMAN.H10MO.B"
TF.rank.plot(motif.pvalue = TF.meth.cor, 
             motif = motif,
             save = FALSE) 
```

## TF ranking plot:  sub-family classification
Shown are TF ranking plots based on the score ($-log_{10}(P value)$) of association between TF expression and DNA methylation of the P73 motif in the LUSC cancer type. The dashed line indicates the boundary of the top 5% association score. The top 3 associated TFs and the TF sub-family members=(dots in red) that are associated with that specific motif are labeled in the plot.

```{r,eval=TRUE,fig.cap="Shown are TF ranking plots based on the score (-log(P value)) of association between TF expression and DNA methylation of the SOX2 motif in the LUSC cancer type. The dashed line indicates the boundary of the top 5% association score. The top 3 associated TFs and the TF sub-family members=(dots in red) that are associated with that specific motif are labeled in the plot"}
load("result/getTF.hypo.TFs.with.motif.pvalue.rda")
motif <- "SOX2_HUMAN.H10MO.B"
TF.rank.plot(motif.pvalue = TF.meth.cor,  
            motif = motif, 
            TF.label = createMotifRelevantTfs("subfamily")[motif],
            save = FALSE)
```

